---
title: "Setting Up a Data Analysis Server in the VT Cloud"
author:
- Tom Woteki, Statistics, drwo@vt.edu, (THW)
- Bob Settlage, Advanced Research Computing, rsettlag@vt.edu, (RES)
output:
  pdf_document: default
  html_document:
    df_print: paged
  html_notebook: default
---
```{r include = FALSE}
knitr::opts_chunk$set(eval=FALSE)
```

# Introduction
This notebook documents the steps for setting up a powerful R-based data analysis server on a virtual machine (VM) hosted in VT's on-premises cloud or  on a late model, multi-core Mac laptop or desktop or equivalent. Once you have completed these steps you will have a VM running the latest versions of R and RStudio Server plus the packages needed for data wrangling, plotting and deep learning, and access to all the statistical analysis packages that are available under R. 

The procedures provided here were tested for both of these configurations:

1) On a VM running Ubuntu 18.04 hosted by VMWare running on a late model multi-core Mac Pro laptop.
2) On a "med" size VM running Ubuntu 18.04 in the VT cloud. The procedures should work on any of the available sizes of VMs running 18.04, although we have not tested them on smaller servers.

For installation on a VM in the VT cloud you will need an account provided by VT's Advanced Research Computing (ARC) team. Installation of VMWare, which is licensed software, and installation of an Ubuntu VM hosted by VMWare are not covered here [^1]. 

[^1]:By default VMWare  allocates 1 gigabyte of memory and 1 core to the VM. For the installation described here we allocated 2 cores and 8 gigabytes of memory to the VM. One gigabyte of memory is not sufficient to compile some of the packages required for a successful installation. Less than 8 gigabytes of memory may be sufficient but we have not tested that.

This guide was developed from scripts and discussions available in the public domain. References are provided below. In the section headings of some sections we have indicated which references were the primary sources.

If you are reading this guide as a pdf you can pull the original R Notebook from which it was rendered at Woteki's git repository: https://github.com/drwo/RSS. We recommend you download it and import it into RStudio on your local machine and conveniently perform the installation step by step from there. If you have any suggestions or find any errors you can push your comments back to git for incorporation into a later version.

## Components and Order of Installation
The following are the components and the order in which they are installed:

1) The latest version of R
2) The latest open source version of RStudio Server (RSS) [^2]. 
3) Several very useful R packages including tidyverse, which comprises a number of useful packages including dplyr and ggplot2; shiny, for building shiny apps; keras for deep learning modeling; and the packages necessary to knit rmarkdown documents to pdf.
4) MySQL, the open source relational database

[^2]:The commercial version of RStudio Server is licensed.

## Installing From the "bash" Shell
Almost all of the installation process is accomplished using bash shell commands in a terminal window connected to the VM. You should have some familiarity with the bash shell, especially since most of the commands are executed with "root" privileges, meaning that you have unrestricted control over the VM and can render it inoperable. Root privileges are required for most of the installation steps, for example installing various Linux libraries [^3].

[^3]:Should you render the VM inoperable, such as erroneously removing a key system directory making it impossible to reboot the machine, the easiest thing to do would be to discard the VM and start over with a new one.

Prior to performing the installation steps you should ssh into the destination VM from the terminal within RStudio. We recommend running each command separately, rather than running  entire chunks as a whole so that you can see the result of each command more easily. Either use the Cmd-Option-Enter key combination on a command selection or cut and paste the code to the terminal window.

## Installing R Packages
As mentioned above, a number of useful R packages are installed as part of this process. These installations are done from the R console using the install.packages function. We have specified the library path for installation so that these packages are installed in a site-wide library and available to all users by default. 

## Preliminaries: Setting up an SSH Server - VMWare VM Only
If you are performing this installation on a VM hosted under VMWare you need to install the ssh service before you can ssh into the server.

First log onto the VM directly from VMWare, then execute these commands from its Terminal app:
```{bash}
sudo apt-get install openssh-server
```

Now go to the Settings app on the VM and find the IP address of the server. Then proceed to ssh into the VM from the terminal window:
```{bash}
ssh <your_user_name>@<your_vm_ip_address>
```


# 1. Installing R (References 1 and 2)

The following commands install the Ubuntu packages for R. The commands mirror those found in references 1 and 2. First we need to add the source of the packages to the sources list. We also want to make sure we get the most recent R, which should be 3.5 as of March 19, 2019.

```{bash}
# Prior to any of this you must have logged into the VM via ssh in your terminal window
# su to root. If you are running this on a VM in the VT cloud this requires 2 factor authentication
sudo su -

apt-get update -y
# Install the packages necessary to add a new repository over HTTPS:
apt install -y apt-transport-https software-properties-common
# Enable the CRAN repository and add the CRAN GPG key to the system
apt-key adv --keyserver keyserver.ubuntu.com --recv-keys E298A3A825C0D65DFD57CBB651716619E084DAB9
add-apt-repository 'deb https://cloud.r-project.org/bin/linux/ubuntu bionic-cran35/'
# Now that the repository is added, update the packages list and install the R package
# Answer yes to any questions concerning installation of components
apt update
apt upgrade
apt-get install r-base r-base-core r-recommended -y
# exit root for now and remain in the terminal window
exit
```

Check to see what version of R is running and do a little computing. It should be the latest version, which as of 3/19/19 is 3.5.x.
```{bash}
R -e "v<-rep(pi,4)+6;v"
```

# 2. Install 64-bit RStudio Server (RSS)
At this point R is installed. If you are still in the R shell quit it.

## 2.1 Create Users & Group
We are going to set up a group "rstudio" and create two user accounts, "rstudio" and "ruser":

1) The group rstudio will define which users can log into RSS. That is, users will have to be members of this group in order to log into RSS.
2) The user rstudio will be the account you use to administer RSS, specifically to add R packages to a site-wide library so that they are accessible to all users [^4]. This user will be in the group rstudio. You should pick a strong password for this user.
3) The user called ruser, with password "ruser", will be a member of the rstudio group. You can use this user name to allow anyone to log into RSS and take it for a test drive.

[^4]: You can always log in with root privileges to add packages to the site-wide library. Doing so as user rstudio will make it more convenient to do so.

In addition to the above you need to add yourself to the rstudio group so that you can log into RSS.

First set up the group and user "rstudio". Log in as root again:
```{bash}
sudo su -
apt-get install gdebi-core -y
groupadd rstudio
# in response to the adduser command set a strong password for user rstudio.
adduser -ingroup rstudio rstudio
# add user rstudio to the sudo group. This gives user rstudio root privileges.
usermod -aG sudo rstudio
```

Now set up the user "ruser" with password "ruser":
```{bash}
adduser -ingroup rstudio ruser
```

Be sure to add yourself to the rstudio group: usermod -aG rstudio <your_user_name>.

## 2.2 Install RStudio Server
Now install RStudio Server. You should be logged in as root.
```{bash}
# you should be logged in as root
cd ~
wget https://download2.rstudio.org/rstudio-server-1.1.463-amd64.deb
gdebi rstudio-server-1.1.463-amd64.deb
```

Verify the RSS installation:
```{bash}
# this should return no output
rstudio-server verify-installation
# this should tell you that RSS is running
rstudio-server status
```

## 2.3 Edit rserver.conf
Here you modify /etc/rstudio/rserver.conf so that one must be a member of the group rstudio in order to log into RSS. As noted above, you should have added yourself to this group.

```{bash}
# you should still be logged in as root
# make /etc/rstudio/rserver.conf writeable
chmod 777 /etc/rstudio/rserver.conf
# insert the line that will require membership in group rstudio
echo "auth-required-user-group=rstudio" >> /etc/rstudio/rserver.conf
# return /etc/rstudio/rserver.conf to its original permissions
chmod 755 /etc/rstudio/rserver.conf
# restart RSS is running so it gets the new conf
#systemctl status rstudio-server.service
systemctl restart rstudio-server.service
```

RSS is now up and running and completely configured. Test logging into RSS as yourself and the other two users. Enter http://<your_VM_ip_address_here>:8787

## 2.4 Firewalls and Ports 22 and 8787

There are 2 firewalls that control access to ports on your VM. Ports are the entry point to services such as RSS; RSS listens on port 8787 by default. Your VM comes equipped with a firewall called ufw that you can optionally activate as described here. There is also a firewall on the VT network that likewise controls access to services on your VM. In order for you to log into RSS on a VT VM, port 8787 must be open on the VM and on the network. The following is the default situation depending on where your VM is hosted:

1) Whether your VM is running under VMWare hosted on a local machine or running in the VT cloud, ufw is inactive and all ports are open on the VM upon installation.
2) If your VM is in the VT cloud ports 22, 8787, 3306 and 7077, plus several others, are open on the network by default [^5].
3) In either case you have been able to log in via ssh and you can log into RSS at port 8787 without taking further action.

Therefore, if you choose to activate the firewall on the VM you need to open ports 22 (for ssh) and 8787 (for RSS) and any other ports that are required to complete this installation, such as 3306 for installing MySQL, which we do below.

[^5]: MySQL listens on 3306. We install MySQL below. Spark clusters need port 7077. Spark is not covered here.

The following commands will activate the firewall and open ports 22 and 8787. If you do activate the VM's firewall be certain to open port 22. Otherwise you will not be able to log back into the VM via a terminal window even though the port is open on the network.

First open port 22. You can safely disregard the message that doing so may disrupt services and answer yes to proceed.
```{bash}
# Yous should still be logged in as root. If not: sudo su -
ufw status verbose
ufw enable
ufw allow 22
```

Now open 8787 and verify the open ports:
```{bash}
ufw allow 8787
ufw status verbose
```

## 2.5 Completing the Installation Process
Now that you have successfully installed RSS you should log into it on your VM using http://<your_VMs_ip_address>:8787 to complete the rest of the installation process. You should log into the server as user rstudio so that the packages we install below are installed in a site-wide library. All the packages will then be automatically available to any user. If you log in with your user name the package installation procedures will fail.

After logging into RSS as user rstudio, the R console and the Terminal window will both be connected to the VM; all R and shell commands will be executed on the VM. After you log in you should upload this file using the upload button on the Files tab. This will make it very convenient to complete the installation process.

# 3. Install Selected R Packages
This section installs some selected R packages into a site-wide library. You should be logged into RSS as user rstudio and have uploaded this notebook using the Files tab.

## 3.1 Setting Up for System-Wide Installation
This section sets up RSS so that user rstudio can install packages into a site-wide library making any packages there available to all users. The following chunk displays the site-wide libraries that are available:

```{r}
# you should be working in the R console window
.libPaths()
```

The first library directory listed is exclusive to user rstudio. We are going to modify /usr/local/lib/R/site-library so that user rstudio can install packages there.
```{bash}
# you should be in the terminal window
cd /usr/local/lib/R
ll
```
Change ownership of the directory site-library to user rstudio:
```{bash}
sudo chown rstudio:rstudio ./site-library
ll
```

## 3.2 Utility Installation Function
The following utility function installs packages into the site-wide library directory now owned by rstudio. It takes as its argument a character vector of packages to install. Run this chunk from the R console to create the function.

```{r}
# you should be in the R console window
load_packages <- function(my_packages) {
  
  #local function to check if a library is installed, if not, install, if so, load
  load_a_package <- function(pack_name) {
    # pack_name is the name of a package to install
    if(!(pack_name %in% installed.packages())) {
      install.packages(pack_name, lib = "/usr/local/lib/R/site-library")
    }
    # require will return true if the library is loaded
    return(require(pack_name, character.only=TRUE))
  }
  
  sapply(my_packages, load_a_package)  
}
```


## 3.1 Install Rcpp
The first package to install is Rcpp which is needed to compile packages that are written in C++, such as dplyr. Upon successful installation the function will return TRUE and Rcpp will be visible and loaded in the Packages window.

```{r echo = FALSE}
load_packages("Rcpp")
```


## 3.2 Install "tidyverse"
"tidyverse" is an extremely useful package. We highly recommend that it be a standard part of an R installation. The tidyverse comprises many other packages: broom, cli, crayon, dbplyr, dplyr, forcats, ggplot2, haven, hms, httr, jsonlite, lubridate, magrittr, modelr, pillar, purrr, readr, readxl, reprex, rlang, rstudioapi, rvest, stringr, tibble, tidyr, xml2, tidyverse.

See: https://tidyverse.tidyverse.org/reference/tidyverse_packages.html

Before installing tidyverse we need to get the following Linux packages: libssl-dev, libxml2-dev, libcurl4-openssl-dev. Switch to the terminal window.
```{bash}
# you need  be in the terminal window
# log in as root
sudo su -
apt-get install libssl-dev
apt-get install libxml2-dev
apt-get install libcurl4-openssl-dev
exit
```

We are now ready to do a clean install of tidyverse. The installation is time consuming and the output is voluminous. Switch to the R console. 
```{r echo = FALSE}
tidyverse_packs <- load_packages("tidyverse")
tidyverse_packs
```

## 3.3 Install Additional Pacakges
These are packages that are useful in many circumstances:

1) "rmarkdown" is required for marking up notebooks. 
2) "knitr", "caTools", "bitops", "rprojroot" enable knitting notebooks to PDFs.
3) "shiny" is for creating shiny apps. Installing it throws a lot of warnings regarding deprecated declarations but the installation seems to conclude successfully. However, we have not tested this.
4) "data.table" is a useful alternative to some portions of the tidyverse packages for wrangling large datasets
5) "ggmap" and "maptools" are handy for developing map graphics and visualizations. Installation of maptools throws some warnings but appears to install successfully but we have not yet tested it.

Before installing these R packages we install the following Linux libraries that are required for knitting to PDFs.
The 2nd command, downloading fonts and other items, may take 10 to 15 minutes. Return to the terminal window.
```{bash}
sudo su -
apt-get -y install texinfo
apt-get -y install texlive-fonts-extra
apt-get -y install pandoc
exit
```

Now install the list of R packages. This will take a while. Return to the R console.
```{r message = FALSE}
xtra_packages <- c("caTools", "bitops", "rprojroot", "rmarkdown", "knitr", "shiny", "data.table",  "ggmap", "maptools", "ineq")
xtra_packs <- load_packages(xtra_packages)
xtra_packs
```

# 4. MySQL Installation (Ref 5)
MySQL as an open source relational database.  Here we install MySQL and administrator accounts with full privileges.

Return to the terminal window.
```{bash}
# login as root
sudo su -
# update apt package index and upgrade
apt update
apt upgrade
apt install mysql-server
systemctl status mysql
```
Now that mysql is running we secure it and set up administrator accounts. As matter of preference, we are not going to configure the VALIDATE_PASSWORD_PLUGIN, which is not as secure as using it. We also choose to:

1) Remove anonymous users
2) Disallow root to log in remotely
3) Remove test databases tables
4) Reload privileges

```{bash}
# secure the installation with a strong password with options as above and exit
mysql_secure_installation
```
Now log into mysql as root.
```{bash}
mysql
```
Now change the authentication method from auth_socket to mysql_native_password using the password you just set up. See Ref 5.
```{mysql}
ALTER USER 'root'@'localhost' IDENTIFIED WITH mysql_native_password BY 'your_strong_password_goes_here';
FLUSH PRIVILEGES;
```
Exit mysql and log back in as root.
```{mysql}
exit
mysql -u root -p
```

Now set up an administrator account, mysql, with root type privileges.
```{mysql}
GRANT ALL PRIVILEGES ON *.* TO 'mysql'@'localhost' IDENTIFIED BY 'MySQL!Warrior' WITH GRANT OPTION;
```
Exit mysql and log in as the user mysql:
```{bash}
exit
mysql -u mysql -p
```
## 4.2 Creating and Additional Admin Account
Note that the admin account we set up, "mysql", is for logging in to mysql on the VM only. To set up a user that can log in from anywhere with all privileges, execute the following commands logged in as the mysql user.

CREATE USER 'some_user'@'%' IDENTIFIED BY 'users_strong_password';
GRANT ALL PRIVILEGES ON *.* TO 'some_user'@'%' IDENTIFIED BY 'users_strong_password' WITH GRANT OPTION;

## 4.1 Accessing MySQL Remotely
MySQL listens on port 3306. If you have enabled the firewall on your VM you need to open the port. You should still be logged in as root. If not, do so.
```{bash}
ufw allow 3306
ufw status verbose
```

NOTE: The default installation of MySQL only allows logging into the server from the VM. To enable remote login you must edit the file /etc/mysql/mysql.conf.d/mysqld.cnf changing the value of the variable "bind-address" from 127.0.0.1 to 0.0.0.0. You can do that by uploading the file mysqld.cnf included here and replacing the existing file on the VM. We do not cover that here.

## 4.2 Package RMySQL
The R package RMySQL is needed to access the MySQL server from an R program. We also need another Linux library. You should still be logged into the VM as root. If not do so.
```{bash}
apt-get install libmariadbclient-dev
exit
```
Return to the R console and install RMySQL:
```{r}
load_packages("RMySQL")
```

# References
The following articles are the sources we consulted to develop this installation kit. There is a great deal of overlap among these guides but also some differences. The installation process tested here is a composite of information from these references and steps that Bob Settlage and I previously tried out.

1) https://linuxize.com/post/how-to-install-r-on-ubuntu-18-04/
2) https://rtask.thinkr.fr/blog/installation-of-r-3-5-on-ubuntu-18-04-lts-and-tips-for-spatial-packages/
3) https://cran.r-project.org/bin/Linux/ubuntu/
4) https://www.edgarsdatalab.com/2017/11/24/setup-an-rstudio-server-in-ubuntu/
5) https://linuxize.com/post/how-to-install-mysql-on-ubuntu-18-04/

